Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 setup.tex
spell-checked first rough draft

\subsubsection{Vertically isothermal disks}
For the special case $\Gamma=1$ the disk is isothermal in the
vertical direction. The equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ is then 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks for simplicitly.

@@

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ amd $\nu$ being
the real frequency and growth rate, respectively.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial lengthscales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all cofficients in
the above equations at the fiducial radius $r=r_0$. However, the full
vertical dependence of the coefficients are retained. Then we can
assume perturbations have spatial dependence in the form
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r},
\end{align}
and we assume $|k_rr|\gg1$ so that curvature terms can also be
ignored.

>>>>

\subsubsection{Vertically isothermal disks}
For the special case $\Gamma=1$ the disk is isothermal in the
vertical direction. The equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ is then 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks for simplicity.

@@

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ and $\nu$ being
the real frequency and growth rate, respectively.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial lengthscales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all coefficients in
the above equations at the fiducial radius $r=r_0$. However, the full
vertical dependence of the coefficients are retained. Then we can
assume perturbations have spatial dependence in the form
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r},
\end{align}
and we assume $|k_rr|\gg1$ so that curvature terms can also be
ignored.

### 2015-02-10 summary.tex
updating
first rough draft
spell-checked first rough draft

\section{Summary and discussion}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega(r,z)$ of a baroclinic disk. The VSI was analysed and
explained earlier in the local limit by \cite{urpin98} and
\cite{urpin03}; and simulated by \cite{arlt04} and more recently by
\cite{nelson13}.

@@

In accordance with the above studies, which show that the VSI applies
to perturbations with small radial lengthscales, we applied the 
radially-local approximation in the linearized fluid equations. For
analytic discussion only, we further simplied the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because previous works show that rapid thermal
relaxation ($\beta\ll1$) is generally required for the VSI. However,
the VGSB framework is not valid in the adiabatic limit.

@@

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we accounted for
the background radial disk structure for consistency. We focused
on the fundamental mode to compare with our analytical
discussion, and because \cite{nelson13} found that this mode 
eventually dominate their simualtions. Our numerical solutions are
mostly consistent with our analytic expectations, and we verified the
dependence of $\beta_\mathrm{crit}$ on disk parameters.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

\emph{Radial dependence.} We have adopted the radially-local approximation 
throughout this paper. This is based on previous works, including
nonlinear simulations, which show that the VSI is relevant for
perturbations with small radial lengthscales. Although this simplifies
the linear problem to ordinary differential equations, this approach
introduces the wavenumber $k_x$ which is a free parameter. The effect
of radial boundary conditions cannot be explored. Future
generalizations of the linear problem should retain the radial
dependence properly and solve the two-dimensional $(r,z)$ partial 
differential equation.

>>>>

\section{Summary and discussion}\label{summary}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega(r,z)$ of a baroclinic disk. The VSI was analyzed and
explained earlier in the local limit by \cite{urpin98} and
\cite{urpin03}; and simulated by \cite{arlt04} and more recently by
\cite{nelson13}.

@@

In accordance with the above studies, which show that the VSI applies
to perturbations with small radial lengthscales, we applied the 
radially-local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because previous works show that rapid thermal
relaxation ($\beta\ll1$) is generally required for the VSI. However,
the VGSB framework is not valid in the adiabatic limit.

@@

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we accounted for
the background radial disk structure for consistency. We focused
on the fundamental mode to compare with our analytical
discussion, and because \cite{nelson13} found that this mode 
eventually dominate their simulations. Our numerical solutions are
mostly consistent with our analytic expectations, and we verified the
dependence of $\beta_\mathrm{crit}$ on disk parameters.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can significantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

\emph{Radial dependence.} We have adopted the radially-local approximation 
throughout this paper. This is based on previous works, including
nonlinear simulations, which show that the VSI is relevant for
perturbations with small radial lengthscales. Although this simplifies
the linear problem to ordinary differential equations, this approach
introduces the wavenumber $k_x$ as a free parameter. The effect
of radial boundary conditions cannot be explored. Future
generalizations of the linear problem should retain the radial
dependence properly and solve the two-dimensional $(r,z)$ partial 
differential equation.
